---
title: "Treg neighbourhood analysis"
output: html_notebook
---

#written by Dr. Harikesh S. Wong in R version 3.5.2 
#2020-07-15: this script determines the local clustering and phenotype of Tregs around individual activated Tconv cells and the local clustering of activated Tconv around other activated Tconv. A sample data set is included

#load packages and read in relevant .csv files
```{r}
library(ggplot2)
library(spatstat)
library(viridis)
library(tidyverse)

#import sample data taken from a female 6 week old WT NOD mouse
WT_CD4 <- read.csv("CD4 Tconvs.csv")
WT_Tregs <- read.csv("Tregs.csv")
WT_PD1 <- filter(WT_CD4, PD1>180)        #filter for cells expressing high PD-1
```

#create .pp3 spatial point patterns and calculate distances
```{r}
WT_Tregs.pp3 <- pp3(WT_Tregs[,6],WT_Tregs[,7],WT_Tregs[,8],range(WT_Tregs[,6]),range(WT_Tregs[,7]),range(WT_Tregs[,8]),
                    marks=WT_Tregs$CD25) 
WT_PD1.pp3 <- pp3(WT_PD1[,6],WT_PD1[,7],WT_PD1[,8],range(WT_Tregs[,6]),range(WT_Tregs[,7]),range(WT_Tregs[,8]),
                  marks=WT_PD1$CD25) 

#Calculate 3D cross distances between each CD4+ PD-1+ Tconv and every Treg 
WT_dist <- (crossdist.pp3(WT_PD1.pp3,WT_Tregs.pp3))%>% 
  t()%>%
  as.data.frame()

#Calculate 3D pairwise distances between every CD4+ PD-1+ Tconv  
WT_PD1_pd <- pairdist.pp3(WT_PD1.pp3)

cutoff <- 50 #user selected parameter - distance cutoff in micrometers

WT_dist_list <- list()

#for each CD4+ PD-1+ Tconv, determine whether each Treg falls within distance cutoff
i=1
for(i in 1:length(WT_dist)) {
  WT_dist_list[[i]] <- which(WT_dist[,i]<= cutoff)
}

#for each CD4+ PD-1+ Tconv, determine the total number of Tregs and other CD4+ PD-1+ Tconvs within the distance cutoff
WC_WT <- as.data.frame(apply(t(WT_dist), 1, function(x) {sum(x<=cutoff)}))
  colnames(WC_WT) <- c("#Tregs_within_cutoff")

WC_WT_PD1 <- as.data.frame(apply((WT_PD1_pd), 1, function(x) {sum(x<=cutoff)}))
  colnames(WC_WT_PD1) <- c("PD1")
```

#extract the measured parameters from the Tregs that fall within the cutoff
```{r}
#add an indexing column to identify invidual Tregs
WT_Tregs <- mutate(WT_Tregs,
                   index = seq(from=1,to=nrow(WT_Tregs),by=1)
)

WT_nr <- list()

i=1
for(i in 1:length(WT_dist)) {
  WT_nr[[i]] <- as.data.frame(filter(WT_Tregs,WT_Tregs$index %in% WT_dist_list[[i]])) 
}

#For each CD4+ PD1+ Tconv, calculate the median parameter values of surrounding Tregs within the distance cutoff
WT_nr_summary <- lapply(WT_nr, function(x) apply(subset(x, select = c(pSTAT5, Foxp3,CD25,PD1)),2, FUN=median))

WT_nr_res <- data.frame()
i=1
for(i in 1:length(WT_dist)) {
  WT_nr_res <- rbind(WT_nr_res,t(WT_nr_summary[[i]]))
}

#calculate the IL-2RA (CD25) expression ratio between each individual CD4+ PD1+ Tconv and surrounding Tregs within distance cutoff
WT_nr_res <- WT_nr_res%>%
 mutate(Tconv_CD25 = WT_PD1$CD25,
        TTconv_Treg_ratio = Tconv_CD25/CD25)
```

#plot data using ggplot
```{r}
#summary
WT_overall <- as.data.frame(cbind(WC_WT$`#Tregs_within_cutoff`,
                                  WC_WT_PD1$PD1,WT_nr_res$TTconv_Treg_ratio))
colnames(WT_overall) <-c("Tregs","CD4PD1", "Tconv_Treg_CD25")

ggplot(WT_overall)+
  stat_density2d(aes(x=Tregs,y=CD4PD1,colour=stat(level)),alpha=0.9,h=c(12,12))+
  scale_color_viridis(option="C",alpha=0.9)+ #include limits to standardize the scale across conditions
  ylim(-10,100)+xlim(0,160)+coord_fixed()+
  theme_bw()+theme(legend.position="none",axis.ticks.length=unit(0.3,"cm"))

ggplot(WT_overall)+
  ylim(-10,100)+xlim(0,160)+coord_fixed()+
  geom_point(aes(x=Tregs,y=CD4PD1,colour=Tconv_Treg_CD25),size=2.5)+
  scale_color_viridis(limits=c(0,1.5),alpha=0.9)+
  theme_bw()+theme(legend.position="none",axis.ticks.length=unit(0.3,"cm"))
```

